Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Guideline for International Students

### 2024-01-09 demystify.tex
updates

\documentclass[oneside,11pt]{memoir}
\setsecnumdepth{subsection}
\setcounter{tocdepth}{3}
\usepackage[T1]{fontenc}
\usepackage[margin=1.5in]{geometry}
\usepackage{soul}
% \usepackage[small,compact]{titlesec} %very powerful
\usepackage[most]{tcolorbox}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=blue,
  urlcolor=blue,
}
\usepackage{booktabs}

@@

\title{Demystifying the Computer Science PhD Admission in the US\\{\large A Guide for International Students}}

@@

\chapterstyle{pedersen}

\makeatletter
\def\maketitle{%
  \null
  \thispagestyle{empty}%
  \vfill
  \begin{center}\leavevmode
    \normalfont
    {\LARGE\raggedright \textbf{\@title}\par}%
    \vfill%
    % \hrulefill\par
    {\Large \@author\par}%
    \vfill%
    {\large\raggedleft \@date\par}%
  \end{center}%
  \vfill
  \null
  \cleardoublepage
}
\makeatother
% \author{ThanhVu H. Nguyen}
% \title{TITLE OF BOOK}
% \date{}

@@

\begin{center}
\url{nguyenthanhvuh.github.io/phd-cs-us/demystify.pdf}, 
\end{center}

@@

\begin{enumerate}
  \item Should you apply to a CS PhD in the US?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\S\ref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically three faculty (\S\ref{sec:evalapps}).
          \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so contact and introduce yourself (\S\ref{sec:contact}).
        \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, personal statement, and recommendation letters.
          \item LORs are very important, but only if they are personalized and talk about your research ability (\S\ref{sec:lor}).
          \item Personal statement is very important. Write it in such a way that makes you \emph{stand out} (\S\ref{sec:research-statement} and \S\ref{sec:improve-your-chance})
          \item GRE \emph{is not} required (\S\ref{sec:grades}). Spend your time on something else.
          \item Grades are important, but depend on the reputation of your school (\S\ref{sec:grades}).
          \item Getting an interview is typically a \emph{good sign} (no one interviews weak candidates).
        \end{itemize}
  \item What to do after getting admitted?
        \begin{itemize}
          \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\S\ref{sec:accepted}).
        \end{itemize}
  \item Funding
        \begin{itemize}
          \item TA and RA are two main sources of funding.  TA (teaching assistantship) is provided by the department to help profs. with classes (e.g., grading). RA (research assistantship) is provided by profs. to help them with their research (\S\ref{sec:funding}).
        \end{itemize}
  \item Choosing School and Professors
        \begin{itemize}
          \item Many schools do not offer PhD studies in CS and many CS professors \emph{cannot} advise or graduate PhD students  (\S\ref{sec:schoolsandprofs}).
          \item Contacting a prof. for research opportunities is recommended, but do it \emph{properly} (\S\ref{sec:contact}).
        \end{itemize}
  \item Miscs and FAQS
        \begin{itemize}
          \item Increasing your admission chance by being unique and standing out (\S\ref{sec:improve-your-chance}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\S\ref{sec:non-stem}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\S\ref{sec:non-us-differences}).
          \item Your PhD costs about \$400K in total, but you \emph{do not} pay for it (\S\ref{sec:ra-cost}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\S\ref{sec:happy}).
        \end{itemize}
\end{enumerate}

@@

%\paragraph{Communicating with LoR writers and People Who Support You}

%stressfull time, rely on support from others
%ask ppl to review your SOPs
%provide information for LORs, ..
%

@@

\begin{table}
  \centering
  \small
  \caption{Top 50 CS PhD programs in the U.S. (\href{https://csrankings.org}{CSRankings}, Jan. 2024). \red{$^*$} indicates that the university has \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Vietnamese prof.} that can advise CS PhD students.}\label{tab:ranking}
  \begin{tabular}{rl|rl}
    \toprule
    1 & Carnegie Mellon & 26 & Duke University \\
    2 & Univ. of Illinois at Urbana-Champaign\red{$^*$}  & 27 & Univ. of California - Santa Barbara \\  
    3 & Univ. of California-San Diego & 28 & Rutgers University\red{$^*$} \\
    4 & Georgia Institute of Technology & 29 & Univ. of California - Riverside\\
    5 & MIT                            & 30 & Pennsylvania State University  \\
    6 & Univ. of California - Berkeley& 31 & Northwestern University\\
    7 & University of Michigan - Ann Arbor\red{$^*$}   & 32& George Mason University\red{$^*$}\\  
    8 & University of Washington      &33 &  Texas A\&M University\red{$^*$} \\
    9 &  Stanford University  &34&  University of Utah \\
    10 & Cornell University  & 35 &  North Carolina State University\\\  
    11 & University of Maryland - College Park &  36& Ohio State \\ 
    12 & Northeastern University\red{$^*$} &37& University of Virginia  \\ 
    13 & Purdue University &38& Yale University \\ 
    14 & University of Wisconsin - Madison\red{$^*$} &39 & Univ. of California - Santa Cruz \\
    15 & University of Texas at Austin &40& Brown University \\ 
    16 & University of Pennsylvania &41 & Harvard University \\
    17 & Columbia University \red{$^*$} &42 & Boston University  \\
    18 & Princeton University  & 43& Rice University\\
    19 & New York University  & 44&  University at Buffalo\red{$^*$}\\
    20 &  University of Massachusetts-Amherst\red{$^*$} &45& University of Colorado-Boulder \\
    21 & Univ. of California - Los Angeles &46& University of Illinois at Chicago  \\
    22 & University of Southern California &47& University of Minnesota \\ 
    23 & University of Chicago &48& University of North Carolina\red{$^*$} \\
    24 & Stony Brook University\red{$^*$} &49& Arizona State University\red{$^*$} \\
    25 &  Univ. of California - Irvine&50& Univ. Of California - Davis \\
    \bottomrule
  \end{tabular}
\end{table}
%Virginia Tech\red{$^*$}  \\
Tab.~\ref{tab:ranking} lists the top 50 CS programs in the US from \href{https://www.csrankings.org}{CSRankings.org}, a ranking system  based on faculty publications at top CS conferences.

@@

%\chapter{Additional Resources}

@@

% Prospective PhD students are often unaware of the importance of research experience in their application.  This provides both concrete research experiences and potential LOR writers that can talk about your research ability.
% %This is especially true for students from small colleges where research opportunities are limited.  
% This section aims to provide some guidance on how to gain research experience as an undergraduate student. %This section is written together Didier, a PhD student at GMU who did his undergraduate study in Rwanda.

@@

%1. I'm fine if I extend an offer and you request a delay, even if it's because you're exploring other programs. No issue here, it's similar to job offerings. Of course, I may not be able to wait indefinitely and may extend the offer to another student, but I'll let you know that.
%2. I'm also okay if you find a program or professor that's a better fit. Similar to the first point, no hard feelings at all. However, I'd appreciate a prompt notification of your decision so I can extend an offer to another student.
%3. I'm NOT okay if you accept my offer but then, upon finding something better, tell me of your decision to withdraw Just like in job offers, once you are committed to an offer, there should be a compelling reason to withdraw (and claiming "I found somewhere better" is the weakest excuse).
%While I won't take any actions against you (a company might), I naturally will remember this and may be cautious in future interactions. I'll also reconsider making similar offers to students in similar situations.

>>>>

\documentclass[oneside,11pt]{memoir}
\chapterstyle{pedersen}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1.5in]{geometry}
\usepackage{soul}
% \usepackage[small,compact]{titlesec} %very powerful
\usepackage[most]{tcolorbox}
\setsecnumdepth{subsection}
\setcounter{tocdepth}{3}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float
\usepackage[svgnames]{xcolor}
\usepackage{hyperref}
\hypersetup{
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=blue,
  urlcolor=blue,
}
\usepackage{booktabs}

@@

\title{Demystifying the Computer Science \\
  PhD Admission in the US\\{\large A Guideline for International Students}}

@@

\makeatletter
\def\maketitle{%
  \null
  \thispagestyle{empty}%
  \vfill
  \begin{center}\leavevmode
    \normalfont
    {\LARGE\raggedright \textbf{\@title}\par}%
    \vfill%
    % \hrulefill\par
    {\Large \@author\par}%
    \vfill%
    {\large\raggedleft \@date\par}%
  \end{center}%
  \vfill
  \null
  \cleardoublepage
}
\makeatother

@@

\begin{center}
  \href{https://nguyenthanhvuh.github.io/phd-cs-us/demystify.pdf}{nguyenthanhvuh.github.io/phd-cs-us/demystify.pdf},
\end{center}

@@

\begin{enumerate}
  \item Should you apply?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\S\ref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically three faculty (\S\ref{sec:evalapps}).
          \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so contact and introduce yourself (\S\ref{sec:contact}).
        \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, personal statement, and recommendation letters.
          \item LORs are very important, but only if they are personalized and talk about your research ability (\S\ref{sec:lor}).
          \item Personal statement is very important. Write it in such a way that makes you \emph{stand out} (\S\ref{sec:research-statement} and \S\ref{sec:improve-your-chance})
          \item GRE \emph{is not} required (\S\ref{sec:grades}). Spend your time on something else.
          \item Grades are important, but depend on the reputation of your school (\S\ref{sec:grades}).
          \item Getting an interview is typically a \emph{good sign} (no one interviews weak candidates).
        \end{itemize}
  \item What to do after getting admitted?
        \begin{itemize}
                \item  Celebrating! Now it is your turn to evaluate the school!
          \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\S\ref{sec:accepted}).
        \end{itemize}
  \item Funding
        \begin{itemize}
          \item TA and RA are two main funding sources.  TA (teaching assistantship) is provided by the department to help profs. with classes (e.g., grading). RA (research assistantship) is provided by profs. to help them with their research (\S\ref{sec:funding}).
        \end{itemize}
  \item Choosing School and Professors
        \begin{itemize}
          \item Many schools do not offer PhD studies in CS and many CS professors \emph{cannot} advise or graduate PhD students  (\S\ref{sec:schoolsandprofs}).
          \item Contacting a prof. for research opportunities is recommended, but do it \emph{properly} (\S\ref{sec:contact}).
        \end{itemize}
  \item Miscs and FAQS
        \begin{itemize}
          \item Increasing your admission chance by being unique and standing out (\S\ref{sec:improve-your-chance}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\S\ref{sec:non-stem}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\S\ref{sec:non-us-differences}).
          \item Your PhD costs about \$400K in total, but you \emph{do not} pay for it (\S\ref{sec:ra-cost}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\S\ref{sec:happy}).
        \end{itemize}
\end{enumerate}

@@

% \paragraph{Communicating with LoR writers and People Who Support You}

% stressfull time, rely on support from others
% ask ppl to review your SOPs
% provide information for LORs, ..
%

@@

\begin{table}
  \centering
  \small
  \caption{Top 50 CS PhD programs in the U.S. (\href{https://csrankings.org}{CSRankings}, Jan. 2024). \red{$^*$} indicates that the university has \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Vietnamese prof.} that can advise CS PhD students.}\label{tab:ranking}
  \begin{tabular}{rl|rl}
    \toprule
    1 & Carnegie Mellon & 26 & Duke University \\
    2 & Univ. of Illinois at Urbana-Champaign\red{$^*$}  & 27 & Univ. of California - Santa Barbara \\  
    3 & Univ. of California-San Diego & 28 & Rutgers University\red{$^*$} \\
    4 & Georgia Institute of Technology & 29 & Univ. of California - Riverside\\
    5 & MIT                            & 30 & Pennsylvania State University  \\
    6 & Univ. of California - Berkeley& 31 & Northwestern University\\
    7 & University of Michigan - Ann Arbor\red{$^*$}   & 32& George Mason University\red{$^*$}\\  
    8 & University of Washington      &33 &  Texas A\&M University\red{$^*$} \\
    9 &  Stanford University  &34&  University of Utah \\
    10 & Cornell University  & 35 &  North Carolina State University\\\  
    11 & University of Maryland - College Park &  36& Ohio State \\ 
    12 & Northeastern University\red{$^*$} &37& University of Virginia  \\ 
    13 & Purdue University &38& Yale University \\ 
    14 & University of Wisconsin - Madison\red{$^*$} &39 & Univ. of California - Santa Cruz \\
    15 & University of Texas at Austin &40& Brown University \\ 
    16 & University of Pennsylvania &41 & Harvard University \\
    17 & Columbia University \red{$^*$} &42 & Boston University  \\
    18 & Princeton University  & 43& Rice University\\
    19 & New York University  & 44&  University at Buffalo\red{$^*$}\\
    20 &  University of Massachusetts-Amherst\red{$^*$} &45& University of Colorado-Boulder \\
    21 & Univ. of California - Los Angeles &46& University of Illinois at Chicago  \\
    22 & University of Southern California &47& University of Minnesota \\ 
    23 & University of Chicago &48& University of North Carolina\red{$^*$} \\
    24 & Stony Brook University\red{$^*$} &49& Arizona State University\red{$^*$} \\
    25 &  Univ. of California - Irvine&50& Univ. Of California - Davis \\
    \bottomrule
  \end{tabular}
\end{table}
% Virginia Tech\red{$^*$}  \\
Tab.~\ref{tab:ranking} lists the top 50 CS programs in the US from \href{https://www.csrankings.org}{CSRankings.org}, a ranking system  based on faculty publications at top CS conferences.

@@

% \chapter{Additional Resources}

@@

% Prospective PhD students are often unaware of the importance of research experience in their application.  This provides both concrete research experiences and potential LOR writers that can talk about your research ability.
% % This is especially true for students from small colleges where research opportunities are limited.
% This section aims to provide some guidance on how to gain research experience as an undergraduate student. %This section is written together Didier, a PhD student at GMU who did his undergraduate study in Rwanda.

@@

% 1. I'm fine if I extend an offer and you request a delay, even if it's because you're exploring other programs. No issue here, it's similar to job offerings. Of course, I may not be able to wait indefinitely and may extend the offer to another student, but I'll let you know that.
% 2. I'm also okay if you find a program or professor that's a better fit. Similar to the first point, no hard feelings at all. However, I'd appreciate a prompt notification of your decision so I can extend an offer to another student.
% 3. I'm NOT okay if you accept my offer but then, upon finding something better, tell me of your decision to withdraw Just like in job offers, once you are committed to an offer, there should be a compelling reason to withdraw (and claiming "I found somewhere better" is the weakest excuse).
% While I won't take any actions against you (a company might), I naturally will remember this and may be cautious in future interactions. I'll also reconsider making similar offers to students in similar situations.
